Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-09 paper.tex
minor

Although the problem is NP-hard in general, fast algorithms are known for specific highly regular lattices, such as the root lattices $A_n$ and $D_n$, their dual lattices $A_n^*$ and $D_n^*$, the integer lattice $\ints^{n}$, and the Leech lattice~\cite[Chap. 4]{SPLAG}\cite{Conway1982FastQuantDec,Conway1986SoftDecLeechGolay, Clarkson1999:Anstar, McKilliam2008, McKilliam2008b, McKilliam2009CoxeterLattices,Vardy1993_leech_lattice_MLD}.  In this paper we consider a particular class of lattices, those of \emph{Voronoi's first kind}~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases,Valentin2003_coverings_tilings_low_dimension,Voronoi1908_main_paper}.  Each lattice of Voronoi's first kind has what is called an \emph{obtuse superbasis}.  We show that if the obtuse superbasis is known, then a closest lattice point can be computed in $O(n^4)$ operations.  This is achieved by enumerating a series of \emph{relevant vectors} of the lattice.  Each relevant vector in the series can be computed in $O(n^3)$ operations using an algorithm for computing a minimum cut in an undirected flow network.  We show that the series converges to a closest lattice point after at most $n$ terms, resulting in $O(n^4)$ operations in total.  This result extends upon a recent result by some of the authors showing that a short vector in a lattice of Voronoi's first kind can be found by computing a minimum cut in a weighted graph~\cite{McKilliam_short_vectors_first_type_isit_2012}.

@@

Let $\Lambda$ be a lattice in $\reals^m$ and let $y \in \reals^m$. A simple method to compute the lattice point $x \in \Lambda$ closest to $y$ is as follows.  Let $x_0$ be some lattice point from $\Lambda$, for example the origin.  Consider the following iteration,
\begin{align}\
x_{k+1} &= x_k + v_k \nonumber \\
v_k &= \arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|, \label{eq:relvectsminimslicer}
\end{align} 
where $\relevant(\Lambda) \cup \{\zerobf\}$ is the set of relevant vectors of $\Lambda$ including the origin.  Observe that the minimum over $\relevant(\Lambda) \cup \{\zerobf\}$ may not be unique.  That is, there may be multiple vectors from $\relevant(\Lambda) \cup \{\zerobf\}$ that are closest to $y - x_k$.  In this case, any one of the minimisers may be chosen.  The results that we will describe do not depend on this choice. We make the following straightforward propositions.

@@

 \begin{proposition}\label{obs:2}
 There is a finite number $K$ such that $x_K, x_{K+1}, x_{K+2}, \dots$ are all closest points to $y$.
 \end{proposition}
 \begin{proof}
Suppose no such finite $K$ exists, then
\[
\|y - x_0\| >  \|y - x_1\| > \|y - x_2\| > \dots
\]
and so $x_0,x_1,\dots$ is an infinite sequence of distinct (due to the strict inequality) lattice points  all contained inside an $n$-dimensional hypersphere of radius $r = \|y - x_0\|$ centered at $y$.  This is a contradiction since, if $\rho$ is the packing radius of the lattice, then less than $\ceil{r/\rho}^n$ lattice points lie inside this sphere by Proposition~\ref{eq:latticepointsinsphere}. 
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors and so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is typically the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory, and so the algorithm can require an amount of memory that grows exponentially with $n$.  We will show that for a lattice of Voronoi's first kind the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

The other factor affecting the complexity is the number of iterations required before the algorithm arrives at a closest lattice point, that is, the size of $K$.  Propositon~\ref{eq:latticepointsinsphere} suggests that this number can be as large as $\ceil{r/\rho}^n$ where $r = \|y - x_0\|^2$ and $\rho$ is the packing radius of the lattice.  Thus, the number of iterations required can potentially grow exponentially with $n$.  The number of iterations required depends on the lattice point that starts the iteration $x_0$.  It is helpful for $x_0$ to be, in some sense, a close approximation of the closest lattice point $x_K$.  Unfortunately, computing close approximations of a closest lattice point is known to be computationally difficult~\cite{feige_inapproximability_2004}.  We will show that for a lattice of Voronoi's first kind a simple and easy to compute choice for $x_0$ ensures that a closest lattice point is reached in at most $n$ iterations, and so $K \leq n$.  Combining this with the fact that each iteration of the algorithm requires $O(n^3)$ operations results in an algorithm that requires $O(n^4)$ operations to compute a closest point in a lattice of Voronoi's first kind.

@@

\begin{theorem} \label{thm:revvecssuperbase} (Conway and Sloane~\cite[Theorem~3]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases})
The relevant vectors of $\Lambda$ are of the form,
\[
\sum_{i \in I} b_i
\]
where $I$ is a strict subset of $\{1, 2, \dots, n+1\}$ that is not empty, i.e. $I \subset \{1, 2, \dots, n+1\}$ and $I \neq \emptyset$.
\end{theorem}

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
We will use the iterative procedure described in~\eqref{eq:relvectsminimslicer} to do this.  In what follows we will assume that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the projection of $y$ orthogonal into the space spanned by $b_1,\dots,b_{n}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + p_k \nonumber \\
p_k &= \arg\min_{p \in \{0,1\}^{n+1}}\| B(z - u_k - p) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  Observe that the procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ be a subset of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with columns given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j} \leq 0$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

We denote by $\min(p)$ and $\max(p)$ the minimum and maximum values obtained by the elements of the vector $p$.  We define the function
\[
\rng(p) = \max(p) - \min(p)
\] 
to return the difference between the maximum and minimum of $p$.  For example, if $p = (2,-1,4) \in \ints^3$ then $\rng(p) = 4 - (-1)=5$.  We might also write simply $\rng(4,-1,5)=5$.  Observe that $\rng(p)$ cannot be negative and that if $\rng(p) = 0$ then all of the elements of $p$ are equal.  We define the function $\subrng(p)$ to return the largest subset, say $S$, of the indices of $p$ such that $\min\{p_i, i \in S\} - \max\{p_i, i \notin S\} \geq 2.$  If no such subset exists then $\subrng(p)$ is the empty set $\emptyset$.  For example, 
\[
\subrng(2,-1,4) = \{1,3\}, \qquad \subrng(2,1,3) = \emptyset, \quad  \subrng(1,3,1) = \{2\}.
\]  
To make the definition of $\subrng$ clear we give the following alternative and equivalent definition.  Let $p \in \ints^n$ and let $\sigma$ be the permutation of the indices $\{1,\dots,n\}$ that puts the elements of $p$ in ascending order, that is
\[
p_{\sigma(1)} \leq p_{\sigma(2)} \leq \dots \leq p_{\sigma(n)}.
\]  
Let $T$ be the smallest integer from $\{2,\dots,n\}$ such that $p_{\sigma(T)} - p_{\sigma(T-1)} \geq 2$.  If no such integer $T$ exists then $\subrng(p) = \emptyset$.  Otherwise 
\[
\subrng(p) =  \{ \sigma(T), \sigma(T+1), \dots, \sigma(n) \}.
\]
The following straightforward property of $\subrng$ will be useful.

@@

\begin{lemma}\label{lem:decrngpreservesclosestpoints}
If $v \in \ints^{n+1}$ such that $B(\floor{z} + v)$ is a closest lattice point to $y = Bz$, then $B\big(\floor{z} + \decrng(v)\big)$ is also a closest lattice point to $y$.
\end{lemma}
\begin{proof}
The lemma is trivial if $\subrng(v) = \emptyset$ so that $\decrng(v) = v$.  It remains to prove the lemma when $\subrng(v) \neq \emptyset$.  In this case put $S = \subrng(v)$ and put $u = \decrng(v) = v - I_S$.  Let $\zeta = z - \floor{z}$ be the column vector containing the fractional parts of the elements of $z$.  We have $\zeta - u = \zeta - v + I_S$.  Applying part~\ref{eq:lem:decSellingsinc} of Lemma~\ref{lem:decSellings} with $p = \zeta - v$ we obtain
\begin{align}
\|B(\zeta - v)\|^2 - \|B(\zeta - u)\|^2 &= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i - v_i)  - 2(\zeta_j - v_j) \big) \nonumber \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j)\big). \label{eq:sumsumBzBu}
\end{align}
Observe that $\zeta_i =  z_i - \floor{z_i} \in [0,1)$ for all $i=1,\dots,n+1$ and so $-1 < \zeta_i-\zeta_j < 1$ for all $i,j=1,\dots,n+1$.  Also, for $i \in S$ and $j\notin S$ we have 
\[
v_i - v_j \geq \min\{ v_i, i \in S\} - \max\{v_j, j \notin S\} \geq 2
\]
by definition of $\subrng(v) = S$.  Thus,
\[
1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j) < 1 + 2 - 4 = -1 < 0.
\]
Substituting this inequality into~\eqref{eq:sumsumBzBu} and using that $q_{ij} \leq 0$ (the obtuse condition~\eqref{eq:obtusecond}) we find that
\[
\|B(z - \floor{z} - v)\|^2 - \|B(z - \floor{z} - u)\|^2 \geq 0,
\]
and so $B(\floor{z} + u) = B\big(\floor{z} + \decrng(v)\big)$ is a closest lattice point to $y = Bz$ whenever $B(\floor{z} + v)$ is.
\end{proof}

>>>>

Although the problem is NP-hard in general, fast algorithms are known for specific highly regular lattices, such as the root lattices $A_n$ and $D_n$, their dual lattices $A_n^*$ and $D_n^*$, the integer lattice $\ints^{n}$, and the Leech lattice~\cite[Chap. 4]{SPLAG}\cite{Conway1982FastQuantDec,Conway1986SoftDecLeechGolay, Clarkson1999:Anstar, McKilliam2008, McKilliam2008b, McKilliam2009CoxeterLattices,Vardy1993_leech_lattice_MLD}.  In this paper we consider a particular class of lattices, those of \emph{Voronoi's first kind}~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases,Valentin2003_coverings_tilings_low_dimension,Voronoi1908_main_paper}.  Each lattice of Voronoi's first kind has what is called an \emph{obtuse superbasis}.  We show that if the obtuse superbasis is known, then a closest lattice point can be computed in $O(n^4)$ operations.  This is achieved by enumerating a series of \emph{relevant vectors} of the lattice.  Each relevant vector in the series can be computed in $O(n^3)$ operations using an algorithm for computing a minimum cut in an undirected flow network~\cite{Picard_min_cuts_1974,Sankaran_solving_CDMA_mincut_1998,Ulukus_cdma_mincut_1998,Cormen2001}.  We show that the series converges to a closest lattice point after at most $n$ terms, resulting in $O(n^4)$ operations in total.  This result extends upon a recent result by some of the authors showing that a short vector in a lattice of Voronoi's first kind can be found by computing a minimum cut in a weighted graph~\cite{McKilliam_short_vectors_first_type_isit_2012}.

@@

Let $\Lambda$ be a lattice in $\reals^m$ and let $y \in \reals^m$. A simple method to compute the lattice point $x \in \Lambda$ closest to $y$ is as follows.  Let $x_0$ be some lattice point from $\Lambda$, for example the origin.  Consider the following iteration,
\begin{align}\
x_{k+1} &= x_k + v_k \nonumber \\
v_k &= \arg\min_{ v \in \relevant(\Lambda) \cup \{\zerobf\} } \|y - x_k - v \|, \label{eq:relvectsminimslicer}
\end{align} 
where $\relevant(\Lambda) \cup \{\zerobf\}$ is the set of relevant vectors of $\Lambda$ including the origin.  The minimum over $\relevant(\Lambda) \cup \{\zerobf\}$ may not be unique, that is, there may be multiple vectors from $\relevant(\Lambda) \cup \{\zerobf\}$ that are closest to $y - x_k$.  In this case, any one of the minimisers may be chosen.  The results that we will describe do not depend on this choice. We make the following straightforward propositions.

@@

 \begin{proposition}\label{obs:2}
 There is a finite number $K$ such that $x_K, x_{K+1}, x_{K+2}, \dots$ are all closest points to $y$.
 \end{proposition}
 \begin{proof}
Suppose no such finite $K$ exists, then
\[
\|y - x_0\| >  \|y - x_1\| > \|y - x_2\| > \dots
\]
and so $x_0,x_1,\dots$ is an infinite sequence of distinct (due to the strict inequality) lattice points  all contained inside an $n$-dimensional hypersphere of radius $r = \|y - x_0\|$ centered at $y$.  This is a contradiction because, if $\rho$ is the packing radius of the lattice, then less than $\ceil{r/\rho}^n$ lattice points lie inside this sphere by Proposition~\ref{eq:latticepointsinsphere}. 
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors and so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is typically the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory and so the algorithm can require an amount of memory that grows exponentially with $n$.  We will show that for a lattice of Voronoi's first kind the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

The other factor affecting the complexity is the number of iterations required before the algorithm arrives at a closest lattice point, that is, the size of $K$.  Propositon~\ref{eq:latticepointsinsphere} suggests that this number might be as large as $\ceil{r/\rho}^n$ where $r = \|y - x_0\|^2$ and $\rho$ is the packing radius of the lattice.  Thus, the number of iterations required might grow exponentially with $n$.  The number of iterations required depends on the lattice point that starts the iteration $x_0$.  It is helpful for $x_0$ to be, in some sense, a close approximation of the closest lattice point $x_K$.  Unfortunately, computing close approximations of a closest lattice point is known to be computationally difficult~\cite{feige_inapproximability_2004}.  We will show that for a lattice of Voronoi's first kind a simple and easy to compute choice for $x_0$ ensures that a closest lattice point is reached in at most $n$ iterations and so $K \leq n$.  Combining this with the fact that each iteration of the algorithm requires $O(n^3)$ operations results in an algorithm that requires $O(n^4)$ operations to compute a closest point in a lattice of Voronoi's first kind.

@@

\begin{theorem} \label{thm:revvecssuperbase} (Conway and Sloane~\cite[Theorem~3]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases})
The relevant vectors of $\Lambda$ are of the form,
\[
\sum_{i \in I} b_i
\]
where $I$ is a strict subset of $\{1, 2, \dots, n+1\}$ that is not empty, i.e., $I \subset \{1, 2, \dots, n+1\}$ and $I \neq \emptyset$.
\end{theorem}

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
The iterative procedure described in~\eqref{eq:relvectsminimslicer} will be used to do this.  In what follows it is assumed that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is a closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the orthogonalprojection of $y$ into the space spanned by $b_1,\dots,b_{n}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + p_k \nonumber \\
p_k &= \arg\min_{p \in \{0,1\}^{n+1}}\| B(z - u_k - p) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  Observe that the procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$, let $S$ be a subset of the indices of $p$, and let $B$ be the matrix from~\eqref{eq:matrxBobtusebasis} with columns given by the superbasis vectors $b_1,\dots,b_{n+1}$.  The following equalities hold:
\begin{enumerate}
\item  ${\displaystyle \|Bp\|^2 - \|B(p - I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i)}$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsinc}
%\item  ${\displaystyle \|Bp\|^2 - \|B(p + I_S + I_T)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)}$, \label{eq:lem:decSellingsincanddec}
\end{enumerate}
where $q_{ij} = \dotprod{b_i}{b_j} \leq 0$ are the Selling parameters.
\end{lemma}
\begin{proof}
We give a proof for part~\ref{eq:lem:decSellingsdec}.  The proof for part~\ref{eq:lem:decSellingsinc} is similar.  Observe that
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
\]
and
\[
\sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
\]
since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
\begin{align*}
\|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
&= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
&= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
\end{align*}
This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - I_S$.  It follows that
\[
\|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
\]
where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
\[
d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
\]
If $i \in S$ and $j \notin S$ then
\[
d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
\]
while if $i \notin S$ and $j \in S$ then
\[
d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
\]
We thus have
\begin{align*}
\|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
\end{align*}
where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

% \begin{lemma}\label{lem:decSellingsincanddec}
% With the same symbols as in Lemma~\ref{lem:decSellings} and with $T$ a subset of the indices of $p$,
% \[
% \|Bp\|^2 - \|B(p + I_S + I_T)\|^2 = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j) + \sum_{i \in T}\sum_{j \notin T}q_{ij}(1 + 2p_i - 2p_j) + \sum_{i \in S}\sum_{j \in T} 2q_{ij}
% \]
% \end{lemma}

We denote by $\min(p)$ and $\max(p)$ the minimum and maximum values obtained by the elements of the vector $p$.  We define the function
\[
\rng(p) = \max(p) - \min(p)
\] 
to return the difference between the maximum and minimum of $p$.  For example, if $p = (2,-1,4) \in \ints^3$ then $\rng(p) = 4 - (-1)=5$.  We might also write simply $\rng(4,-1,5)=5$.  Observe that $\rng(p)$ cannot be negative and that if $\rng(p) = 0$ then all of the elements of $p$ are equal.  We define the function $\subrng(p)$ to return the largest subset, say $S$, of the indices of $p$ such that $\min\{p_i, i \in S\} - \max\{p_i, i \notin S\} \geq 2.$  If no such subset exists then $\subrng(p)$ is the empty set $\emptyset$.  For example, 
\[
\subrng(2,-1,4) = \{1,3\}, \qquad \subrng(2,1,3) = \emptyset, \quad  \subrng(1,3,1) = \{2\}.
\]  
To make the definition of $\subrng$ clear we give the following alternative and equivalent definition.  Let $p \in \reals^n$ and let $\sigma$ be the permutation of the indices $\{1,\dots,n\}$ that puts the elements of $p$ in ascending order, that is
\[
p_{\sigma(1)} \leq p_{\sigma(2)} \leq \dots \leq p_{\sigma(n)}.
\]  
Let $T$ be the smallest integer from $\{2,\dots,n\}$ such that $p_{\sigma(T)} - p_{\sigma(T-1)} \geq 2$.  If no such integer $T$ exists then $\subrng(p) = \emptyset$.  Otherwise 
\[
\subrng(p) =  \{ \sigma(T), \sigma(T+1), \dots, \sigma(n) \}.
\]
The following straightforward property of $\subrng$ will be useful.

@@

\begin{lemma}\label{lem:decrngpreservesclosestpoints}
If $v \in \ints^{n+1}$ such that $B(\floor{z} + v)$ is a closest lattice point to $y = Bz$, then $B\big(\floor{z} + \decrng(v)\big)$ is also a closest lattice point to $y$.
\end{lemma}
\begin{proof}
The lemma is trivial if $\subrng(v) = \emptyset$ so that $\decrng(v) = v$.  It remains to prove the lemma when $\subrng(v) \neq \emptyset$.  In this case put $S = \subrng(v)$ and put $u = \decrng(v) = v - I_S$.  Let $\zeta = z - \floor{z}$ be the column vector containing the fractional parts of the elements of $z$.  We have $\zeta - u = \zeta - v + I_S$.  Applying part~\ref{eq:lem:decSellingsinc} of Lemma~\ref{lem:decSellings} with $p = \zeta - v$ we obtain
\begin{align}
\|B(\zeta - v)\|^2 - \|B(\zeta - u)\|^2 &= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i - v_i)  - 2(\zeta_j - v_j) \big) \nonumber \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j)\big). \label{eq:sumsumBzBu}
\end{align}
Observe that $\zeta_i =  z_i - \floor{z_i} \in [0,1)$ for all $i=1,\dots,n+1$ and so $-1 < \zeta_i-\zeta_j < 1$ for all $i,j=1,\dots,n+1$.  Also, for $i \in S$ and $j\notin S$ we have 
\[
v_i - v_j \geq \min\{ v_i, i \in S\} - \max\{v_j, j \notin S\} \geq 2
\]
by definition of $\subrng(v) = S$.  Thus,
\[
1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j) < 1 + 2 - 4 = -1 < 0 \qquad \text{for $i \in S$ and $j \notin S$}.
\]
Substituting this inequality into~\eqref{eq:sumsumBzBu} and using that $q_{ij} \leq 0$ (the obtuse condition~\eqref{eq:obtusecond}) we find that
\[
\|B(z - \floor{z} - v)\|^2 - \|B(z - \floor{z} - u)\|^2 \geq 0
\]
and so $B(\floor{z} + u) = B\big(\floor{z} + \decrng(v)\big)$ is a closest lattice point to $y = Bz$ whenever $B(\floor{z} + v)$ is.
\end{proof}
